Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-06-15 draft_lm1.tex
fixes #36
Yes it is, now in boldface to make clear. fixes #38
Hopefully closes #35

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the Appendix.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\citep{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\citep{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction factor), which approaches $1$ at convergence~\citep{brooks1998}.

@@

This approach thus makes it possible to incorporate epidemiological knowledge 
about disease Biology (e.g. the magnitude of $R_0$) into the computation of \rr.
This may prove particularly important when disease counts are low and/or close 
to the detection threshold.
We provide an R script to perform the above elicitation at 
\url{https://github.com/fccoelho/paperLM1/blob/master/R/elicit_Rt_prior.R}.

>>>>

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the \textbf{Appendix}.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\citep{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\citep{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection (see 
\textbf{Appendix} for details).
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction factor), which approaches $1$ at convergence~\citep{brooks1998}.

@@

This approach thus makes it possible to incorporate epidemiological knowledge 
about disease Biology (e.g. the magnitude of $R_0$) into the computation of \rr.
This may prove particularly important when disease counts are low and/or close 
to the detection threshold.
We provide an R script to perform the elicitation described above at 
\url{https://github.com/fccoelho/paperLM1/blob/master/R/elicit_Rt_prior.R}.

\subsection*{DiffeRential Evolution Adaptive Metropolis (DREAM)}

Since the posterior distributions desired in this paper are not available in 
closed-form, we need to resort to numerical methods to obtain 
approximations.
We employ an adaptive Markov chain Monte Carlo (MCMC) algorithm, proposed
by~\cite{vrugt2008}, called  DiffeRential Evolution Adaptive Metropolis (DREAM).

DREAM draws on the basic structure of Differential Evolution Markov Chain 
(DE-MC) which runs $N$ independent chains in parallel and accepts moves 
proportional to the difference of two randomly sampled members (chains), thus 
differentially evolving conditional on the proposal variances.
An additional aspect of DREAM is the so-called delayed rejection.
This procedure adapts the original Metropolis-Hastings algorithm by attempting 
new moves whenever a proposal is rejected (instead of setting the current state 
to the previously accepted one), thus delaying the rejection of proposals. 
This potentially decreases autocorrelation and improves mixing 
(see~\cite{mira2001} for details).

A desirable property of DREAM is that one can scale the algorithm with model 
complexity, meaning one can initialise as many parallel chains as there are 
parameters in the model, thus exploiting the multi-core architecture of modern 
computers to speed up convergence of the MCMC.

A Python implementation of DREAM for dynamic models is available in the package 
Bayesian inference with Python (BIP)~\citep{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
The code to produce the results presented in this paper is available 
at~\url{https://github.com/fccoelho/paperLM1/tree/master/python}.
